NanoPDLIM2 enhanced efficacy of PD-1 blockade and chemotherapy in mouse lung cancers

Version 1

Abstract

Abstract

Background

Immune checkpoint inhibitors (ICIs) and their combination with other therapies such as chemotherapy, fail in most cancer patients. We previously identified the PDZ-LIM domain-containing protein 2 (PDLIM2) as a bona fide tumor suppressor that is repressed in lung cancer to drive cancer and its chemo- and immunotherapy resistance, suggesting a new target for lung cancer therapy improvement.

Methods

Human clinical samples and data were used to investigate PDLIM2 genetic and epigenetic changes in lung cancer. Using an endogenous mouse lung cancer model faithfully recapitulating refractory human lung cancer and a clinically feasible nano-delivery system, we investigated the therapeutic efficacy, action mechanism, and safety of systemically administrated Pdlim2 expression plasmids encapsulated in nanoparticles (nanoPDLIM2) and its combination with PD-1 antibody and chemotherapeutic drugs.

Results

PDLIM2 repression in human lung cancer involves both genetic deletion and promoter methylation. NanoPDLIM2 showed low toxicity, high tumor specificity, antitumor activity, and greatly improved the efficacy of anti-PD-1 and chemotherapeutic drugs, with complete tumor remission in 60% of mice and substantial tumor reduction in the remaining mice by the combination of three therapies. Mechanistically, nanoPDLIM2 increased major histocompatibility complex class I (MHC-I) expression, suppressed multi-drug resistance 1 (MDR1) induction, nuclear Rela and stat 3, and survival genes (Bcl-xl and cycline D1) in tumor cells; meanwhile it enhanced lymphocyte tumor infiltration and activation, thus turning the cold tumors hot and sensitive to ICIs and rendering them vulnerable to chemotherapeutic drugs.

Conclusions

These studies established a clinically applicable PDLIM2-based combination therapy with significantly improved efficacy for lung cancer and possibly other cold cancers.

Introduction

Lung cancer is the leading cause of cancer-related deaths in both men and women with a 5-year survival rate of only 22% [1]. Although these outcomes may be improved by immune checkpoint blockade therapy involving the disruption of the binding of programmed cell death 1 (PD-1, also known as CD279) on tumor-infiltrating lymphocytes (TILs) to programmed death receptor ligand 1 (PD-L1, also known as B7-H1 or CD274) on tumor and tumor-associated cells, most lung cancer patients still fail the therapy, with a response rate of only about 20% [2]. In general, this revolutionary immunotherapy works better against ‘hot’ tumors, which have abundant TILs in the tumor microenvironment, strong immunogenicity and sufficient PD-L1 expression on cancer cells. Unfortunately, most tumors are ‘cold’, with low T-cell infiltration, weak immunogenicity and minimal PD-L1 expression, and show weak response to immune checkpoint inhibitors (ICIs) [3, 4].

Cold lung tumors without targetable oncogenic drivers are treated with chemotherapy as the standard approach [5]. However, the response rate to this conventional cancer therapy is also low and resistance often occurs after an initial response, with an overall survival (OS) of about 12-17 months. Given its roles of inducing TILs and immunogenicity, in particular PD-L1 expression, chemotherapy can be an ideal candidate for combination with PD-1/PD-L1 blockade to improve therapeutic efficacy [6]. Indeed, combination treatment with ICIs and chemotherapeutic drugs shows synergy and better efficacy in both preclinical animal models and clinical trials of lung cancer and other cancers [6–8]. However, even with the combination of ICIs and chemotherapy, tumors in animals do not remit completely and the objective response rate (ORR) of lung cancer patients only reaches 33%-49.7%, with a median progression-free survival (PFS) of just 5.1-9 months and a median OS of 13-22 months [7, 8]. Thus, further improvement over the chemo-immunotherapy is direly need.

Our recent studies on human and animal lung cancers have shown that most lung tumors not only have low TILs and decreased PD-L1, but also down-regulate major histocompatibility complex class I (MHC-I), thus evading recognition and attack by CD8+ T cells, including those unleashed by ICIs and/or recruited by chemotherapy [6, 9, 10]. Following our initial studies [11–17], we have established the PDZ-LIM domain-containing protein PDLIM2, also known as SLIM or mystique [18–20], as a bona fide tumor suppressor, and its repression is a causative driver of lung cancer and resistance to ICIs and chemotherapeutic agents [6]. One most important function of PDLIM2 is to promote the ubiquitination and proteasomal degradation of nuclear signal transducer and activator of transcription 3 (STAT3) and nuclear factor-κB (NF-κB) RelA (also known as p65), two master transcription factors that function as proto-oncogenes in lung and many other cancers [6, 10-13, 21-27]. Pdlim2 repression in tumor cells thus results in the persistent activation of Stat3 and Rela, leading to MHC-I downregulation and high expression of tumor growth-related genes [6]. In addition, Pdlim2 deficiency also causes strong induction of multi-drug resistance 1 (MDR1) for acquired chemo-resistance [6]. Therefore, PDLIM2 is a potential target to further enhance the efficacy of chemo-and immune-therapies for lung cancers.

In this study, we explored the mechanism of PDLIM2 repression in human lung cancers and tested whether systemic administration of nanoparticle-encapsulated Pdlim2-expression plasmids (nanoPDLIM2) could enhance the efficacy of anti-PD-1 and/or chemotherapeutic drugs in a mouse lung cancer model. These studies indicate that besides epigenetic repression, loss of heterozygosity (LOH) with copy number losses contributes to PDLIM2 downregulation in about 58% of human lung cancers, and that Pdlim2 heterozygous deletion (Pdlim2+/−) mice develop spontaneous tumors in lung and other organs. Notably, systemic administration of nanoPDLIM2 reverses the phenotypes caused by Pdlim2 repression in lung tumors, and its combination with PD-1 and chemotherapeutic drugs shows the strongest therapeutic efficacy compared to single or double therapies, inducing complete remission of lung tumors in 60% of mice. These findings provide a firm basis to target PDLIM2 to enhance the efficacy of ICIs and chemotherapeutic drugs for lung cancer therapy.

Material and Methods

Animals and lung carcinogenesis

We have complied with all relevant ethical regulations for animal testing and research. The animal experiments were performed in accordance with the US National Institutes of Health (NIH) Guidelines on the Use of Laboratory Animals. All animals were maintained under pathogen-free conditions and used according to protocols approved by the Institutional Animal Care and Use Committee (IACAUC) of the University of Pittsburgh. Pdlim2flx/flx/Spc-rtTAtg/−/(tetO)7CMV-Cretg/tg (ΔSPC) mice under a pure FVB/N background and Pdlim2-/- mice under a pure BALB/c background have been described before [6, 19, 21, 22, 28]. For lung carcinogenesis, six-to eight-week-old ΔSPC and wild type FVB/N mice were intraperitoneally (i.p.) injected with urethane (1 mg/g body weight, Sigma-Aldrich, St. Louis, MO, USA) once a week for six consecutive weeks [6], followed by different treatments as shown in the figures. Mice were sacrificed at six weeks post urethane treatment for examination of lung tumors, inflammation, and treatment-induced toxicity. Surface tumors in mouse lungs were counted blinded under a dissecting microscope, and tumor diameters were measured by microcalipers.

Preparation of Pdlim2-expression plasmid or empty vector plasmid nanoparticles

The polyethylenimine (PEI)-based nanoparticles (in vivo-jetPEI) (Polyplus Transfection, New York, NY, USA) and plasmid DNA complexes at a nitrogen-to-phosphate ratio of 8 (N/P=8) were prepared according to the manufacturer’s instructions. Briefly, 25 µg of pCMV-myc-Pdlim2 [21] or empty vector plasmids in 100 μl of a 5% glucose solution were mixed with the in vivo-jetPEI reagent (4 μl) diluted into 100 μl of a 5% glucose solution. After 15 minutes of incubation at room temperature, the mixed solution (200 μl/mouse) was injected intravenously (i.v.) via the tail vein. Presence of nanoPDLIM2 in tissues were measured by PCR assays targeting Amp-R and Lyz2 in the genome of empty vector and Pdlim2 plasmids.

Histology and immunohistochemistry (IHC) analysis

Lung, liver, kidney, and spleen tissues were excised, fixed in formalin, embedded in paraffin, and cut into 4-μm-thick sections. Sections were stained with H&E, or subjected to sequential incubations with the indicated primary antibodies, biotinylated secondary antibodies and streptavidin-HRP [6]. Antibodies were listed in Supplementary Table S1.

In vivo BrdU labeling

Mice were i.p. injected with 50 mg/kg BrdU (Sigma-Aldrich, St. Louis, MO, USA) 24 h prior to sacrifice. Mouse lung tissue sections were stained with anti-BrdU (Sigma-Aldrich, St. Louis, MO, USA). BrdU labeling index was calculated as the percentage of labeled cells per total cells counted (> 500 cells in each counted tumor-containing area).

Flow cytometry (FACS) analysis

The cells were isolated from mouse lungs using dispase digestion followed filtration with 70 μm Nylon cell strainer; then cells were incubated with the antibodies against cell surface antigens after blocked with αCD16/CD32. The cells were then fixed with paraformaldehyde (2%), permeabilized and incubated with antibodies against intracellular antigens if needed. For interferon-γ (IFNγ) staining, cells were treated with phorbol 12-myristate 13-acetate (PMA, 50 ng/ml), ionomycin (1 μM), brefeldin A (BFA, 3 μg/ml) and monensin (2 μM) for 4 h before they were stained for FACS analysis. Data were acquired and analyzed by Accuri C6 or LSRFortessa I (BD Biosciences) and FlowJo software [29].

Quantitative polymerase chain reaction (qPCR) analysis

The indicated tissues or cells were subjected to DNA or RNA extraction, RNA reverse transcription and real-time PCR using trizol, reverse transcriptase, and Power SYBR Green PCR Master Mix (Thermo Fisher Scientific, Waltham, MA USA) according to manufacturer’s protocol. Primer pairs used for qPCR were listed in Supplementary Table S2.

Microsatellite and gene-specific PCR-based LOH analysis of PDLIM2

Genomic DNAs were isolated from human lung tumors and their matched normal tissues using the PureLink Genomic DNA Purification Kit (Invitrogen, Carlsbad, CA, USA), and subjected to semi- quantitative PCR using the primers specific for the microsatellite markers D8S1786 and D8S1752 that straddle the PDLIM2 genetic locus or the PDLIM2 genetic locus itself. Primer pairs used for the assays were listed in Supplementary Table S2.

Statistical analysis

One-way ANOVA power analysis was used to determine the minimum sample size. Animals were randomly assigned to different treatment groups. Measurements were taken from distinct samples. Student’s t test (two tailed) and one-way ANOVA/Tukey’s or two-way ANOVA/Sidak’s test were used to assess significance of differences between two groups and multiple comparisons, respectively. Pearson’s correlation test was used to assess association between PDLIM2 expression with its promoter methylation or genetic deletion and the overlap between PDLIM2 promoter methylation and genetic deletion. All bars in figures represent means ± standard error of the mean (SEM). The p values are indicated as *p < 0.05, **p < 0.01, ns, not statistically significant, except for those shown in figures. The p values < 0.05 and 0.01 are considered statistically significant and highly statistically significant, respectively. Grubbs’ and ROUT outlier tests were established a priori. There were no exclusions in the analyses.

Results

Both epigenetic repression and genetic deletion of PDLIM2 in lung cancer

Using 40% of the expression level of matched normal lung tissues as the cut-off, PDLIM2 was repressed in over 75% of human lung cancers in a prior study [6]. Analysis of the Cancer Genome Atlas (TCGA) database found that if using 50% as the cut-off, PDLIM2 was repressed in about 93% of human lung cancers (Figure 1A). Further analysis of the TCGA dataset showed that over 70% of human lung cancers with decreased PDLIM2 expression had hypermethylation in PDLIM2 promoter, when using 125% of the methylation level of the PDLIM2 promoter in normal lung tissues as the cut-off (Figure 1A, 1B), validating our previous finding that epigenetic silencing was the main mechanism underlying PDLIM2 repression in lung cancer [6].

PDLIM2 repression in human lung cancers involves both epigenetic alteration and genetic deletion, and PDLIM2 genetic deletion in mice leads to development of spontaneous tumors, majorly lung tumors. (A) TCGA data showing PDLIM2 repression in over 90% of lung cancers if using 50% of the expression level in normal lung tissues as the cut-off (NL, n = 110; Tumor, n = 1019). (B) TCGA data showing PDLIM2 promoter hypermethylation and expression repression (dashed box) in over 70% of lung cancers when using 125% of the methylation level in normal lung tissues as the cut-off (n = 827). (C) TCGA data showing positive associations between PDLIM2 expression and its gene copy numbers as well as PDLIM2 genetic deletion and expression repression (dashed box) in about 58% of lung cancers using the copy number variation of −0.1 as the cut-off (n = 1010). (D) TCGA data showing simultaneous promoter hypermethylation and genomic deletion of PDLIM2 (dashed box) in about 44% of lung cancers (n = 816). (E) Microsatellite-PCR showing pdlim2 loss in human lung cancers (n = 21). (F) qPCR showing PDLIM2 loss in human lung cancer cell lines with known copy number of the PDLIM2 gene (n = 25). (G) Kaplan-Meier tumor-free survival curve showing increased spontaneous tumors in Pdlim2-/- and Pdlim2+/− mice compared to WT mice. Gehan-Breslow-Wilcoxon test was performed. *p < 0.05. (H) Percentage of tumor types spontaneously developed in Pdlim2+/− mice, and 51% of tumors occurred in the lung.

In line with PDLIM2’s location on chromosome 8p21.3, a frequent LOH region in lung and other tumors [30–34], analysis of the TCGA database revealed that PDLIM2 expression was positively associated with its gene copy numbers, and that over 58% of human lung cancers with decreased PDLIM2 expression had genetic deletion of the PDLIM2 gene if copy number variation (CNV) of −0.1 was used as the cut-off for the gene deletion (Figure 1C). Further analysis indicated that about 44% of human lung cancers with decreased PDLIM2 expression simultaneously harbored the promoter hypermethylation and LOH with copy number losses of the PDLIM2 gene, and approximately 26% and 14% of them only having the promoter hypermethylation and LOH, respectively (Figure 1D). The gene deletion of PDLIM2 was confirmed by microsatellite and gene-specific PCR-based LOH analysis of human primary lung cancer tissues and cell lines (Figure 1E, 1F; Supplementary Table S3). These data suggested that PDLIM2 repression in lung cancer involves both epigenetic silencing and genetic deletion.

To determine the significance of PDLIM2 LOH in lung tumor, we examined whether Pdlim2 heterozygous deletion leads to decreased Pdlim2 expression and spontaneous tumors in mice. Unlike its absolute absence in Pdlim2 homozygous deletion (Pdlim2-/-) mice, Pdlim2 was detected in the lung of Pdlim2+/− mice, but at a much lower level compared to wild type (WT) mice (about 50% of WT, Supplementary Figure S1A). Importantly, Pdlim2+/− mice, like Pdlim2-/- mice, also developed spontaneous tumors (Figure 1G). Of note, over 50% of tumors developed in Pdlim2-/- and Pdlim2+/− mice were lung tumors (Figure 1H; Supplementary Figure S1B). These data are also highly consistent with the fact that PDLIM2 is ubiquitously expressed under physiological conditions, with the highest level in the lung and lung epithelial cells in particular [6, 18-20]. Thus, PDLIM2 is a haploinsufficient tumor suppressor that is particularly important for lung tumor suppression.

Efficacy of systemic administration of nanoPDLIM2 in refractory lung cancer

Although reversal of PDLIM2 epigenetic repression by epigenetic drugs to restore PDLIM2 expression in cancer cells may be used to treat lung cancers [6], this approach cannot be applied to lung cancers involving PDLIM2 LOH with copy number losses, as occurred in 58% of all lung cancer cases. To overcome this limitation and expand PDLIM2-targeted therapy to all lung cancers with PDLIM2 repression, we tested the therapeutic efficacy of systemic administration of PDLIM2-expression plasmids encapsulated by the clinically feasible in vivo-jetPEI [35–39]. To this end, we employed mouse lung cancers induced by urethane, a faithful model of human lung cancer and adenocarcinoma (AC) in particular [6, 23, 24, 40-43]. Urethane is a chemical carcinogen present in fermented food, alcoholic beverage and also cigarette smoke, the predominant risk factor accounting for about 90% of human lung cancers [44]. Like its human counterpart, the murine lung cancer induced by urethane also shares Pdlim2 repression, in addition to their similarities in histology, genetics, molecular biology and immunology [6, 9, 23, 24, 40-43]. WT mice with lung tumors induced by urethane were i.v. injected with nanoparticle-encapsulated Pdlim2 plasmids or empty vector plasmids. Six weeks post the initial treatment of nanoPDLIM2, mice were sacrificed and lung tissues were collected for assays (Figure 2A). Compared to the empty vector plasmid mock-treated group (Vec), nanoPDLIM2-treated mice had a significantly reduced tumor burden (−31%) in their lungs, though the response of tumor numbers was weak (Figure 2B).

Systemic administration of PDLIM2 plasmid nanoparticles shows efficacy in mouse model of refractory lung cancer. (A) Schedule of lung cancer induction and treatment. (B) Intravenous administration of PDLIM2 plasmid nanoparticles reduced lung tumors in mouse urethane lung tumor model (n ≥ 6). Nanoparticles carrying an empty vector plasmid (Vec) that was employed to express PDLIM2 were used as a control. (C) IHC staining showing decreased nuclear expression of Stat 3 and Rela in lung tumors by PDLIM2 nanotherapy (n = 6). (D) IHC staining showing decreased Bcl-xl and increased cleaved caspase −3 in lung tumors by PDLIM2 nanotherapy (n = 6). (E) IHC staining showing decreased Cyclin D1 and proliferation (Brdu) in lung tumors by PDLIM2 nanotherapy (n = 6). Scale bar in (C-E), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ± SEM in (B-E). **p < 0.01; ns, not statistically significant.

In line with the reduction in tumor burden, nanoPDLIM2 administration decreased nuclear Rela and Stat3, a hallmark of NF-κB and STAT3 activation, accordingly reduced the expression of their downstream cell survival gene Bcl2l1(Bcl-xl, −49%) and cell proliferation gene Ccnd1 (Cyclin D1, −61%), increased cleaved caspase 3 (+230%), a marker of apoptosis, and decreased proliferation (−47%) of lung cancer cells (Figure 2C-2E). These data indicated that intravenous administration of nanoPDLIM2 reduced lung tumor burden, suppressed oncogenic Rela and Stat3 activation and increased apoptosis of lung tumors.

High tumor specificity and low toxicity of systemic administration of nanoPDLIM2

To characterize the efficiency of systemically administrated nanoPDLIM2, we examined Pdlim2 expression in lung tumors and several organs, including the liver, kidney and spleen. A high level of Pdlim2 was detected in the lung tumors from mice treated with nanoPDLIM2 one week post the injection, whereas no obvious Pdlim2 was detected in the lung tumors from mice of Vec group (Figure 3A, 3B). It should be pointed out that the tumor delivery efficiency of Pdlim2-expression or empty vector plasmid nanoparticles did not differ (Figure 3B). However, either Pdlim2-expression or empty vector plasmids were hardly detected in other organs/tissues of the same mice, including liver, kidney and spleen (Figure 3B). Of note, in the first two days after injection, both empty vector and Pdlim2 plasmids were also detected at high levels in those tissues but quickly declined afterward (data not shown), consistent with the well-documented tumor-specific enrichment of nanoparticles [37, 45-47]. Consistently, ectopic Pdlim2 was not detected, and the levels of Pdlim2 protein were comparable in these tissues (Figure 3B, 3C).

PDLIM2 nanotherapy shows high tumor specificity and low toxicity. (A) IHC staining showing high Pdlim2 re-expression in lung tumors after PDLIM2 nanotherapy. (B) PCR and IB assays showing lung tumor-specific plasmids delivery and PDLIM2 expression by PDLIM2 nanotherapy (n = 3). (C) IHC staining showing comparable expression of PDLIM2 in the indicated tissues of mice treated with PDLIM2 expression plasmid or empty vector plasmid nanoparticles. (D) No significant changes in animal body weight by PDLIM2nanoparticles (n = 5). (E) H&E staining showing no noticeable changes in major organs by PDLIM2nanoparticles nanotherapy. Scale bar: (A and C) 20 μm; (E) 50 μm.

The i.v. injection of either Pdlim2-expression or control plasmid nanoparticles showed no obvious toxicity to animals, as evidenced by no significant changes in body weight and histology of all organs/tissues examined, including the spleen, kidney and liver (Figure 3D, E). Mouse appearance and behaviors, such as eating, drinking, defecating, urinating, sniffing, grooming and digging, were not different between control and nanoparticle groups (data not shown). Taken together, these data suggest nanoPDLIM2 shows therapeutic efficacy in the mouse lung tumor model while causes low toxicity.

Synergy of nanoPDLIM2 with chemotherapy in lung cancer treatment

Given the role of Pdlim2 in inhibiting the expression of cell survival and proliferation genes in tumor cells, which contribute to chemoresistance, we tested whether nanoPDLIM2 increases the efficacy of chemotherapy in the mouse model of lung cancer. Treatment with carboplatin and paclitaxel, two first-line chemotherapeutic drugs for lung and many other cancers, led to a significant decrease in tumor number (−35%) and tumor burden (−87%) (Figure 4A). Importantly, combination with nanoPDLIM2 further decreased both tumor number (−78%) and tumor burden (−94%), suggesting a promising synergy between nanoPDLIM2 and chemotherapy in lung cancer treatment. Consistently, significantly higher tumor cell apoptosis was detected in mice treated with the combination therapy (+650%), in comparison to those with nanoPDLIM2 (+200%) or chemotherapy (+230%) alone (Figure 4B).

PDLIM2 nanotherapy renders lung cancers more vulnerable to chemotherapy. (A) Synergy of PDLIM2 nanotherapy and chemotherapeutic drugs (carboplatin and paclitaxel) to reduce lung tumors in lung tumor models induced by urethane (n ≥ 5). (B) IHC staining showing effects on lung tumor cell apoptosis by PDLIM2nanoparticle, chemo-drugs, and their combination (n = 6). (C) IHC staining showing increase in nuclear Rela by chemotherapy and its inhibition by PDLIM2nanoparticles (n = 6). (D) IHC staining showing MDR1 induction by chemotherapy and its blockage by PDLIM2 nanotherapy (n = 6). Scale bar in (B-D), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ± SEM. *p < 0.05; **p < 0.01; ns, not statistically significant.

Chemotherapy induced strong Rela activation/nuclear expression and Mdr1 expression (Figure 4C, 4D). NanoPDLIM2 not only repressed the constitutive activation of Rela in cancer cells but also prevented the strong induction of Rela activation and Mdr1 expression by the chemotherapy. Thus, nanoPDLIM2 improves the therapeutic efficacy of chemotherapy through blocking both intrinsic and acquired chemoresistance of lung cancer cells.

NanoPDLIM2 enhanced the efficacy of ICIs in lung cancer treatment

High expression of cell survival genes renders tumor cells resistant to the tumoricidal activity of cytotoxic T lymphocytes (CTLs), including those unleashed by ICIs. Given nanoPDLIM2 administration suppressed survival genes in lung tumors (Figure 2E), we hypothesize it enhances the efficacy of immunotherapy. NanoPDLIM2 increased the number of TILs and the expression of MHC-I, the most important and core components of immunotherapies including PD-1 immune checkpoint blockade therapy (Figure 5A, 5B). In line with our previous studies [6], PD-1 blocking antibody reduced tumor burden (−25%) and tumor number (−13%) (Figure 5C). Combining with PDLIM2 nanotherapy further decreased tumor burden (−64%), and caused a trend toward decrease for tumor number. Consistently, higher increases in both CD4+ and CD8+ TILs, stronger CD8+ CTL activation and lung tumor cell death were detected in combination therapies of nanoPDLIM2 and PD-1 antibody (Figure 5D-5F). These data together suggest that nanoPDLIM2 enhanced the therapeutic efficacy of PD-1 blockade.

PDLIM2 nanotherapy increases the efficacy of PD-1 blockade immunotherapy. (A) IHC staining showing increased TILs in lung tumors by PDLIM2nanoparticles (n = 6). (B) FACS showing increased MHC-I expression in lung tumor cells by PDLIM2 nanotherapy (n = 4). (C) Enhancing effects of PDLIM2nanoparticles on PD-1 immunotherapy in lung tumors induced by urethane (n ≥ 5). (D) IHC staining showing increased TILs by PDLIM2 nanotherapy in the context of immunotherapy (n = 6). (E) FACS showing increased activation of CD8+ T cells by PDLIM2 nanotherapy in the context of immunotherapy (n = 3). (F) IHC staining showing increased lung tumor cell apoptosis by PDLIM2 nanotherapy, immunotherapy, and further increase by their combination (n = 6). Scale bar in (A, D, F), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ± SEM. **p < 0.01; ns, not statistically significant.

Complete remission of all lung cancers in most mice by the combination treatment of nanoPDLIM2 and anti-PD-1 and chemotherapeutic drugs

Like most human lung cancers, lung cancers in our animal model exhibit a decreased expression of PD-L1 on cell surface, in additional to the low expression of MHC-I and low number of TILs [6, 9]. In line with our previous finding that PD-L1 expression is largely independent of Pdlim2 [6], nanoPDLIM2 failed to induce PD-L1 expression in lung tumors; in contrast, chemotherapy induced PD-L1 (Figure 6A). This may explain why the enhancing effect of nanoPDLIM2 on PD-1 blockade therapy is only moderate.

Combination of PDLIM2 nanotherapy, chemotherapy and immunotherapy shows enhanced efficacy in lung cancer treatment. (A) IHC staining showing PD-L1 induction by chemotherapy but not by PDLIM2 nanotherapy. (B) Schedule of lung cancer induction and treatment. (C) Urethane-induced lung tumors were more resistant to chemo and αPD-1 combination therapy in mice with Pdlim2 specifically deleted in lung epithelial cells (ΔSPC) (n ≥ 4). (D) IHC staining showing no MHC-I induction by chemotherapy, PD-1 immunotherapy or their combination (n = 6). (E) Tumor examination showing complete remission of all lung tumors in 60% of mice by combination of the three therapies (n ≥ 5). (F) IHC staining showing increased TILs by PDLIM2 nanotherapy in mice treated with anti-PD-1 and chemotherapeutic drugs (n = 6). (G) FACS analysis showing increased lung CD8+ T-cell activation by PDLIM2 nanotherapy in mice treated with anti-PD-1 and chemotherapeutic drugs (n = 5). (H) No significant effect of PDLIM2 nanotherapy on the body weight of mice treated with anti-PD-1 and chemotherapeutic drugs (n = 5). Scale bar in (A and F), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ± SEM in (c-g). *p < 0.05; **p < 0.01; ns, not statistically significant.

Induction of PD-L1 could protect cancer cells from immune attack and thereby restricts further efficacy improvement from its combination with PDLIM2 nanotherapy, which increased TILs (Figure 4A); on the other hand, upregulated PD-L1 may increase the sensitivity of cancer cells to PD-1 blockade therapy. Indeed, combination of chemotherapeutic drugs and PD-1 blockade showed a promising synergy in reducing both tumor number and tumor burdens in comparison to the individual treatment of chemotherapy or PD-1 blockade therapy [6]. The synergy was largely blocked when Pdlim2 was genetically deleted from lung cancer cells (ΔSPC) (Figure 6B, 6C), suggesting an important role of Pdlim2 in the combination therapy of chemotherapeutic drugs and anti-PD-1. Whereas PDLIM2 nanotherapy increased MHC-I expression on tumor cells (Figure 5B), which is critical for better recognition and killing of tumor cells by CD8+ CTLs, chemotherapy alone or its combination with anti-PD-1 failed to do so (Figure 6D). Accordingly, combination of anti-PD-1 and chemotherapeutic drugs, like the combination therapies of nanoPDLIM2 and anti-PD-1 or chemotherapeutic drugs, also failed to induce a complete remission of lung cancers in any mice.

Given their overlapping roles in increasing TILs and in particular their complement roles in inducing MHC-I and PD-L1 expression on tumor cells, which turn tumors hot and more sensitive to PD-1 blockade, better efficacy is expected for combination of nanoPDLIM2, chemotherapeutic drugs, and anti-PD-1 compared to single therapy or combination of double therapies. Indeed, combination of the three showed the strongest effect on reducing tumor burden and numbers, and caused complete remission of lung cancers in 60% of mice (Figure 6E) and much more shrinkage of lung tumors in remaining mice in comparison to those treated with two combined therapies (data not shown). In line with the increase in therapeutic efficacy, the triple combination therapy significantly increased the numbers and/or activation of CD4+ and CD8+ T cells in the lung, compared to the combination of chemotherapeutic drugs and anti-PD-1 (Figure 6F, 6G).

Of note, consistent with the undetectable toxicity of PDLIM2 nanotherapy, its co-treatment did not further increase the toxicity of anti-PD-1 and chemotherapeutic drugs, as evidenced by no obvious histological difference of major organs, including the liver, lung, kidney, and spleen (Supplementary Figure S2), and no significant differences in animal body weights, in comparison to the mice received the combinational treatment of anti-PD-1 and chemotherapeutic drugs in the presence or absence of empty vector plasmid nanoparticles (Figure 6H).

Discussion

PD-1/PD-L1 blockade immunotherapy has recently joined chemotherapy as a standard treatment for lung and several other cancers [2–4]. While some patients have shown dramatic responses, most patients do not benefit from this novel treatment. Currently, various combinations of ICIs with other therapies, in particular with chemotherapeutic drugs, are being extensively tested in both preclinical and clinical trial studies, aiming to expand the benefit of this innovative therapy [6–8]. Although a promising synergy and better efficacy have been observed in some of these studies [6–8], further improvement is direly needed. Using a mouse model of lung cancer, we show, for the first time, that PDLIM2 nanotherapy restored Pdlim2 expression in lung tumors and reduced tumor burden; more importantly, co-treatment of nanoPDLIM2 enhanced the efficacy of anti-PD-1 and chemotherapeutic drugs and induced complete remission of all lung cancers in most animals.

Most human lung cancers as well as lung cancers in our animal model have low numbers of TILs and decreased expression of PD-L1 and MHC-I on the cell surface [6, 9], all of which contribute to the resistance to PD-1 blockade therapy. Through inducing immunogenic cell death (ICD) of cancer cells [8], chemotherapy can increase TILs in tumor microenvironment and PD-L1 expression on tumor cells, and thereby synergize with anti-PD-1. However, chemotherapy cannot induce MHC-I expression, this limits further improvement of its synergy with immune therapy for complete cancer remission. Also, tumor cells usually express high levels of survival genes against the tumorcidal effects of chemotherapeutic drugs and of CTLs, including those activated by chemotherapy and unleashed by ICIs.

On the other hand, nanoPDLIM2 induced MHC-I expression and lymphocyte tumor infiltration but did not up-regulate PD-L1. Moreover, nanoPDLIM2 prevented the induction of Mdr1 and the expression of tumor survival genes, this would further sensitize tumor cells to the cytotoxicity of chemotherapeutic drugs and immune cells. Consistent with these, nanoPDLIM2 improved the efficacy of chemotherapy and PD-1 blockade therapy, and the combination of these three therapies resulted in complete cancer remission in 60% of the animals and dramatic tumor reduction in the remaining mice.

Another important clinical characteristic of PDLIM2 nanotherapy is its tumor-specificity and high safety profile. It delivered Pdlim2-expression plasmids specifically to lung tumor tissues and increased Pdlim2 expression. Of note, nanoPDLIM2 showed undetectable toxicity in the animal model, and did not further increase the toxicity of anti-PD-1 and chemotherapeutic drugs. In contrast, although epigenetic drugs (5-aza-dC and MS-275) could restore PDLIM2 expression in cancer cells with PDLIM2 epigenetic repression [6, 10-15] and showed better efficacy as a monotherapy or combined with chemotherapeutic drugs or anti-PD-1 than nanoPDLIM2 (Supplementary Figure S4), the epi-drugs caused body weight loss and much worse toxicity when combined with chemotherapy (Supplementary Figure S3). In addition, the therapeutic efficiency of epigenetic drugs depends on PDLIM2 expression [6]. Although epigenetic therapy can be applicable to 26% lung cancers with PDLIM2 epigenetic repression, it is not suitable in about 58% of lung cancers harboring PDLIM2 LOH with copy number losses. NanoPDLIM2, on the other hand, provides a strategy to restore PDLIM2 expression in all lung cancers with PDLIM2 repression, regardless of whether it is resulted from promoter hypermethylation or gene deletion.

Conclusions

In summary, the presented data identify genetic deletions as a major mechanism for PDLIM2 repression in human lung cancers, and PDLIM2 as a haploinsufficient tumor suppressor particularly important for suppressing lung cancer and therapy resistance. More importantly, these preclinical data demonstrate that combination treatment of nanoPDLIM2, anti-PD-1 and chemotherapeutic drugs induced complete remission of all lung cancers in most animals with high safety profile, suggesting nanoPDLIM2 as a potential strategy to enhance the efficacy of current therapies.

Abbreviations

adenocarcinoma

brefeldin A

copy number variation

cytotoxic T lymphocyte

fluorescent activated cell sorting

Food and Drug Administration

interferon-γ

immunogenic cell death

immune checkpoint inhibitor

immunohistochemistry

intraperitoneal

intravenous

loss of heterozygosity

multi-drug resistance 1

major histocompatibility complex class I

nanoparticle-encapsulated Pdlim2-expression plasmids

nuclear factor-κB

National Institutes of Health

objective response rate

overall survival

programmed cell death 1

programmed death ligand 1

programmed death ligand 2

PDZ-LIM domain-containing protein 2

polyethylenimine

progression-free survival

phorbol 12-myristate 13-acetate

quantitative polymerase chain reaction

standard error of the mean

signal transducer and activator of transcription 3

The Cancer Genome Atlas

tumor-infiltrating lymphocyte

wild type.

eLife Assessment

The reviewers have found the work to be valuable to the field of immunotherapy in the treatment of cancer. The data supporting the role of PDLIM2 as a tumor suppressor, and more immediately, as a strategy to improve the efficacy of immunotherapy treatment, was viewed as compelling. However, the results are lacking a completed mechanism, which would substantially expand the impact of the work.

Reviewer #1 (Public Review):

The manuscript by Sun and colleagues followed their previous findings on the tumor-suppressive role of PDLIM2 in lung cancer. They further investigated various mechanisms, including epigenetic modification, copy number variation, and LOH, that led to the decreased expression of PDLIM2 in human lung cancer. Next, they used a nanoparticle-based approach to specifically restore the expression in mouse lung tumors. They showed that over-expression of PDLIM2 in lung cancer repressed its progression in vivo. Also, this treatment could synergize with chemotherapy and checkpoint inhibitor anti-PD-1. Overall, the results were quite promising and convincing, using a treatment combination that would appear to have the potential for clinical implementation.

Reviewer #2 (Public Review):

Summary: The authors have previously demonstrated that the E3 ligase PDLIM2 inhibits NF-kB and STAT3 and is epigenetically repressed in human lung cancers (Sun et al. Nat. Comm. 2019 10: 5324); therefore, PDLIM2 is a tumor suppressor in lung cancer. In this manuscript, they follow up on their previous findings and show that expression of PDLIM2 is downregulated in human lung cancers by both genetic deletion and promoter methylation. They further describe a novel approach to restore the expression of PDLIM2 in mouse lung tumors by systemically administering PDLIM2 plasmids encapsulated in nanoparticles (termed "nanoPDLIM2"). The nanoPDLIM2 approach was shown to exhibit efficacy with low toxicity in a urethane-induced mouse lung cancer model. The authors further demonstrated the synergy of nanoPDLIM2 with chemotherapy and PD-1 blockade immunotherapy. The combination therapy of nanoPDLIM2, chemotherapy, and immunotherapy proved most effective with complete tumor remission in 60% of mice. Mechanistically, nanoPDLIM2 upregulated MHC-I expression, enhanced CD4/CD8 T cell activation and tumor infiltration, and suppressed MDR1 induction and nuclear expression of STAT3, RelA and prosurvival genes in tumors. Overall, this study is important because it reinforces the critical roles of PDLIM2 in suppressing lung cancer, and also identifies a potential approach to restoring PDLIM2 expression in lung tumors. The experiments were well executed; the data are convincing and support the conclusions made by the authors.

Reviewer #3 (Public Review):

Strengths:

NanoPDLIM2, nanotechnologies that efficiently deliver lentivirus overcomes resistance to chemotherapy and anti-PD-1 immunotherapy. This is a new strategy for enhancing the efficiency of immune checkpoint inhibitors. This finding is important from a clinical translation perspective, but I have several minor concerns.

Weaknesses:

1. Please describe the mechanism of increased MHC class I and PD-L1 by PDLIM2.

2. Please describe the mechanism of decreased MDR1, nuclear RelA and STAT3 by PDLIM2.

3. Please determine whether PDLIM2 expression directly impacts immune cells (function and number)?

4. What is the efficiency of PDLIM2 delivery? Does delivery efficiency determine anti-tumor effect?

5. Authors used a non-immunogenic tumor model. Can you demonstrate the combination effect with PDLIM2 in immunogenic lung cancer models to determine whether the combination of PDLIM2 with anti-PD-1 Ab confers a synergistic effect without chemotherapy?

6. On page 11, % change can make one over-interpret data.

7. In Figure 5, what is the difference between 5A and 5D?

8. It is unclear whether PDLIM2 confers an additive or a synergistic effect with anti-PD-1/chemo.

9. Have the authors tested any toxicity in normal lungs?

Author Response

Reviewer #3 (Public Review):

Strengths:

NanoPDLIM2, nanotechnologies that efficiently deliver lentivirus overcomes resistance to chemotherapy and anti-PD-1 immunotherapy. This is a new strategy for enhancing the efficiency of immune checkpoint inhibitors.

This finding is important from a clinical translation perspective, but I have several minor concerns.

Weaknesses:

1. Please describe the mechanism of increased MHC class I and PD-L1 by PDLIM2.

Our previous studies showed that PDLIM2 induces MHC-I induction through decreasing STAT3 whereas it is dispensable for PD-L1 expression (Sun et al, 2019, PMID: 31757943). In line with the studies, PD-L1 is induced by chemotherapeutic drugs, but not by NanoPDLIM2 (Figure 6A). Together with the roles of PDLIM2 in repressing RelA-dependent MDR1 induction by chemotherapy and in preventing expression of cell survival and proliferation genes by targeting both RelA and STAT3 (Sun et al, 2019, PMID: 31757943), further providing the mechanistic basis for the combination and synergistic effect of nanoPDLIM2, anti-PD-1 and chemo drugs. The improvement has now been further incorporated.

1. Please describe the mechanism of decreased MDR1, nuclear RelA and STAT3 by PDLIM2.

Our previous studies demonstrated that PDLIM2 reduces MDR1 expression by degrading nuclear RelA (Sun et al, 2019, PMID: 31757943).

1. Please determine whether PDLIM2 expression directly impacts immune cells (function and number)?

As shown in Figure 5, NanoPDLIM2 increased the number and activation of tumor infiltrating lymphocytes (TILs); and in prior study, PDLIM2 knockout repressed the numbers of TILs and inhibited the activation of CD4+ and CD8+ T cells, while its re-expression in lung tumors led to T cell activation (Sun et al. 2019, PMID: 31757943). On the other hand, selective deletion of PDLIM2 in immune cells and in particular myeloid cells repressed the numbers and activation of TILs (Li et al, 2021, PMID: 33539325; PMCID: PMC8021114). Thus, PDLIM2 may impact immune cells both directly and indirectly, particularly when nanoparticles can deliver PDLIM2 into both tumor cells and tumor-associated immune cells (despite PDLIM2 is delivered into much fewer immune cells compared to tumor cells).

1. What is the efficiency of PDLIM2 delivery? Does delivery efficiency determine anti-tumor effect?

As shown in the manuscript, the dose of PDLIM2 used already shows high delivery (20-30 copies per tumor cell in Figure 3B) and therapeutic efficacy in the mouse model of refractory lung cancer and particularly when being combined with anti-PD-1 and chemo drugs. It is of interest to test different doses in the model for the best delivery and efficacy, which is actively being pursued in the lab.

1. Authors used a non-immunogenic tumor model. Can you demonstrate the combination effect with PDLIM2 in immunogenic lung cancer models to determine whether the combination of PDLIM2 with anti-PD-1 Ab confers a synergistic effect without chemotherapy?

Yes, it is of interest to demonstrate the combination of PDLIM2 and anti-PD-1 in immunogenic lung cancer models with chemotherapy although a synergy is highly expected. The greatest challenge in the lung cancer field is the low response of non-immunogenic tumor, which is the focus of the current manuscript.

1. On page 11, % change can make one over-interpret data.

The % change has been removed from the manuscript.

1. In Figure 5, what is the difference between 5A and 5D?

Figure 5A shows the increase of TILs by nanoPDLIM2 in animals that did not receive PD-1 blockade immunotherapy, Figure 5D shows the increase of TILs by nanoPDLIM2 in animals received PD-1 blockade immunotherapy.

1. It is unclear whether PDLIM2 confers an additive or a synergistic effect with anti-PD-1/chemo.

PDLIM2 nanotherapy confers a synergistic effect with chemotherapy on increasing apoptosis in tumors (Figure 4B) and tumor reduction (Figure 4A and 6E, left panel, tumor number), confers a synergistic effect with antiPD-1 on increasing CD4+ and CD8+ TILs (Figure 5A and 5D), and apoptosis in tumors (Figure 5F), and an additive effect on tumor reduction (Figure 5C and 6E), and confers a synergistic effect with chemotherapy plus anti-PD-1 on increasing CD4+ and CD8+ TILs (Figure 5A and 6F) and tumor reduction (Figure 6E, left panel, tumor number).

1. Have the authors tested any toxicity in normal lungs?

Same to tumor lungs, no obvious toxicity has been observed in normal lungs.

Reviewer #1 (Recommendations For The Authors):

The paper is clear and well-written, although some minor edits are needed. For example, the title could be changed to reflect both human and mouse studies in the manuscript for more general readers. Moreover, 'lung cancer' should be used instead of 'lung cancers'. The manuscript could be further improved by validating their findings in a different model and particularly the syngeneic model of metastatic lung cancer for a better overall survival time by the new combination therapy, given the fact that clinical trial studies usually start in patients with metastatic tumors. But this is optional because the therapeutic effect on primary lung cancer is already significant.

Thanks for the correction and wonderful suggestions. The “lung cancers” were replaced with “lung cancer”, and the title was changed to “Improving PD-1 blockade plus chemotherapy for complete remission of lung cancer by nanoPDLIM2”.

Reviewer #2 (Recommendations For The Authors):

1. What is the rationale for i.v. injection of nanoparticles containing PDLIM2 plasmid? Intranasal administration of nanoparticles may potentially target nanoPDLIM2 specifically to the lungs. Another potential option is intranasal infection of mice with adenovirus expressing PDLIM2.

The rationale for i.v. injection of nanoPDLIM2 is that iv injected nanoPDLIM2 first reach into the lung and more importantly tumor tissues as well as the convenience and high efficacy of mouse i.v. injection, particularly when multiple injections are needed. Mice are much less stressful compared to other intranasal or even intratracheal injection. Adenovirus can be used only once, because it will initiate ant-viral immune response in mice.

1. The authors examine PDLIM2 expression in lung tumors 1 week after i.v. administration of nanoparticles (Fig. 3A). Do all tumor cells express PDLIM2 after nanoPDLIM2 treatment? How long does PDLIM2 persist in the tumors? The kinetics of PDLIM2 expression may be informative to help interpret the results from the various combination treatments given to the mice. Multiple rounds of nanoPDLIM2 treatment could potentially improve the efficacy of the treatment.

For all the sections examined (n=6), PDLIM2 was re-expressed in most but not all lung cancer cells at 1-week of the i.v administration. Accordingly, nanoPDLIM2 was injected weekly. We are examining if PDLIM2 reexpression can last longer. We are also testing the best dose with the best efficacy.

1. Does the plasmid DNA from nanoparticles trigger an innate immune response in the lung that contributes to anti-tumor responses?

In line with previous studies showing no effect on immune responses (Bonnet et al. 2008. PMID: 18709489), the dose used in current study does not significantly affect immune cells in the lung, suggesting no obvious effect of nanoparticles with empty plasmid on innate immune response.

1. In Fig. 4, does the combination of nanoPDLIM2 and chemotherapy diminish STAT3 nuclear staining?

NanoPDLIM2 alone decreased nuclear STAT 3 in tumor cells (Figure 2C), it also diminished nuclear STAT3 in tumor cells with the combination of chemotherapy.

Version 2

Abstract

Abstract

Background

Immune checkpoint inhibitors (ICIs) and their combination with other therapies such as chemotherapy, fail in most cancer patients. We previously identified the PDZ-LIM domain-containing protein 2 (PDLIM2) as a bona fide tumor suppressor that is repressed in lung cancer to drive cancer and its chemo and immunotherapy resistance, suggesting a new target for lung cancer therapy improvement.

Methods

Human clinical samples and data were used to investigate PDLIM2 genetic and epigenetic changes in lung cancer. Using an endogenous mouse lung cancer model faithfully recapitulating refractory human lung cancer and a clinically feasible nano-delivery system, we investigated the therapeutic efficacy, action mechanism, and safety of systemically administrated PDLIM2 expression plasmids encapsulated in nanoparticles (nanoPDLIM2) and its combination with PD-1 antibody and chemotherapeutic drugs.

Results

PDLIM2 repression in human lung cancer involves both genetic deletion and epigenetic alteration. NanoPDLIM2 showed low toxicity, high tumor specificity, antitumor activity, and greatly improved the efficacy of anti-PD-1 and chemotherapeutic drugs, with complete tumor remission in most mice and substantial tumor reduction in the remaining mice by their triple combination. Mechanistically, nanoPDLIM2 increased major histocompatibility complex class I (MHC-I) expression, suppressed multi-drug resistance 1 (MDR1) induction and survival genes and other tumor-related genes expression in tumor cells, and enhanced lymphocyte tumor infiltration, turning the cold tumors hot and sensitive to ICIs and rendering them vulnerable to chemotherapeutic drugs and activated tumor-infiltrating lymphocytes (TILs) including those unleashed by ICIs.

Conclusions

These studies established a clinically applicable PDLIM2-based combination therapy with great efficacy for lung cancer and possibly other cold cancers.

Introduction

Lung cancer is the leading cause of cancer-related deaths in both men and women with a 5-year survival rate of only 22% [1]. Although these outcomes may be improved by immune checkpoint blockade therapy involving the disruption of the binding of programmed cell death 1 (PD-1, also known as CD279) on tumor-infiltrating lymphocytes (TILs) to programmed death receptor ligand 1 (PD-L1, also known as B7-H1 or CD274) on tumor and tumor-associated cells, most lung cancer patients still fail the therapy, with a response rate of only about 20% [2]. In general, this revolutionary immunotherapy works better against ‘hot’ tumors, which have abundant TILs, strong immunogenicity and sufficient PD-L1 expression. Unfortunately, most tumors are ‘cold’, with low T-cell infiltration, weak immunogenicity and minimal PD-L1 expression, and show weak response to immune checkpoint inhibitors (ICIs) [3, 4].

Cold lung tumors without targetable oncogenic drivers are treated with chemotherapy as the standard approach [5]. However, the response rate to this conventional cancer therapy is also low and resistance often occurs after an initial response, with an overall survival (OS) of about 12-17 months. Given its roles in inducing TILs and immunogenicity, in particular PD-L1 expression, to turn cold tumors hot, chemotherapy can be an ideal candidate for combination with PD-1/PD-L1 blockade to improve therapeutic efficacy [6]. Indeed, combination treatment with ICIs and chemotherapeutic drugs shows synergy and better efficacy in both preclinical animal models and clinical trials of lung cancer and other cancers [6–8]. However, even with the combination of ICIs and chemotherapy, tumors in animals do not remit completely and the objective response rate (ORR) of lung cancer patients only reaches 33%-49.7%, with a median progression-free survival (PFS) of just 5.1-9 months and a median OS of 13-22 months [7, 8]. Thus, further improvement over the chemo-immunotherapy is direly needed.

Our recent human and mouse studies have shown that most lung tumors not only have low TILs and decreased PD-L1, but also down-regulate major histocompatibility complex class I (MHC-I), evading recognition and attack by CD8+ T cells, including those unleashed by ICIs and/or recruited by chemotherapy [6, 9, 10]. Following our previous cell line studies [11–17], we have established the PDZ-LIM domain-containing protein PDLIM2, also known as SLIM or mystique [18–20], as a bona fide tumor suppressor and its repression as a causative driver of lung cancer and resistance to ICIs and chemotherapeutic agents [6]. While PDLIM2 is epigenetically repressed in human lung cancer, associating with therapeutic resistance and poor prognosis, its global or lung epithelial-specific deletion in mice leads to lung cancer development, chemoresistance, and complete resistance to anti-PD-1 and epigenetic drugs. One most important function of PDLIM2 is to promote the ubiquitination and proteasomal degradation of nuclear signal transducer and activator of transcription 3 (STAT3) and nuclear factor-κB (NF-κB) RelA (also known as p65), two master transcription factors that function as proto-oncogenes in lung and many other cancers [6, 10–13, 21–27]. PDLIM2 repression in tumor cells thus results in the persistent activation of STAT3 and RelA, leading to MHC-I downregulation and high expression of tumor growth-related genes [6]. It also leads to strong induction of multi-drug resistance 1 (MDR1) for acquired chemo-resistance, as chemotherapy further enhances RelA activation for MDR1 transcription in PDLIM2 deficient tumor cells [6].

In this study, we examined whether and how PDLIM2 can be targeted to treat lung cancer in a faithful mouse model of human lung cancer. In particular, we tested whether systemic administration of nanoparticle-encapsulated PDLIM2-expression plasmids (nanoPDLIM2) could enhance the efficacy of anti-PD-1 and/or chemotherapeutic drugs. We also examined whether PDLIM2 repression in human lung cancer involves genetic deletion and its relationship with epigenetic silencing. These studies indicate that besides epigenetic repression, loss of heterozygosity (LOH) contributes to PDLIM2 downregulation in about 58% of human lung tumors, and that Pdlim2 heterozygous deletion (Pdlim2+/-) mice develop spontaneous tumors in lung and other organs. Notably, systemic administration of nanoPDLIM2 reverses the phenotypes caused by PDLIM2 repression and induces complete remission of all lung tumors in most mice without further increasing toxicity when combined with anti-PD-1 and chemotherapeutic drugs. These findings provide a firm basis to combine ICIs and chemotherapeutic drugs with PDLIM2-targeted therapy for the treatment of lung and other cancers.

Material and Methods

Animals and lung carcinogenesis

We have complied with all relevant ethical regulations for animal testing and research. The animal experiments were performed in accordance with the US National Institutes of Health (NIH) Guidelines on the Use of Laboratory Animals. All animals were maintained under pathogen-free conditions and used according to protocols approved by the Institutional Animal Care and Use Committee (IACAUC) of the University of Pittsburgh. Pdlim2flx/flx/SP-C-rtTAtg/−/(tetO)7CMV-Cretg/tg (ΔSPC) mice under a pure FVB/N background and Pdlim2-/- mice under a pure BALB/c background have been described before [6, 19, 21, 22, 28]. For lung carcinogenesis, six-to eight-week-old ΔSPC and wild type FVB/N mice were intraperitoneally (i.p.) injected with urethane (1 mg/g body weight, Sigma-Aldrich, St. Louis, MO, USA) once a week for six consecutive weeks [6], followed by different treatments as shown in the figures. Mice were euthanized at six weeks post urethane treatment for examination of lung tumors, inflammation, and treatment-induced toxicity. Surface tumors in mouse lungs were counted blinded under a dissecting microscope, and tumor diameters were measured by microcalipers.

Preparation of PDLIM2-expression plasmid or empty vector plasmid nanoparticles

The polyethylenimine (PEI)-based nanoparticles (in vivo-jetPEI) (Polyplus Transfection, New York, NY, USA) and plasmid DNA complexes at a nitrogen-to-phosphate ratio of 8 (N/P=8) were prepared according to the manufacturer’s instructions. Briefly, 25Lµg of pCMV-myc-Pdlim2 or empty vector plasmids in 100Lμl of a 5% glucose solution were mixed with the in vivo-jetPEI reagent (4Lμl) diluted into 100Lμl of a 5% glucose solution. After 15Lminutes of incubation at room temperature, the mixed solution (200 μl/mouse) was injected intravenously (i.v.) via the tail vein. PDLIM2-expression plasmid or empty vector plasmid DNA in tissues were measured by PCR assays targeting Amp-R in the genome of these plasmids.

Histology and immunohistochemistry (IHC) analysis

Lung, liver, kidney, and spleen tissues were excised, fixed in formalin, embedded in paraffin, and cut into 4-μm-thick sections. Sections were stained with H&E or subjected to sequential incubations with the indicated primary antibodies, biotinylated secondary antibodies and streptavidin-HRP [6]. Antibodies were listed in Supplementary Table S1.

In vivo BrdU labeling

Mice were i.p. injected with 50Lmg/kg BrdU (Sigma-Aldrich, St. Louis, MO, USA) 24Lh prior to euthanasia. Mouse lung tissue sections were stained with anti-BrdU (Sigma-Aldrich, St. Louis, MO, USA). BrdU labeling index was calculated as the percentage of labeled cells per total cells counted (> 500 cells in each counted tumor-containing area).

Flow cytometry (FACS) analysis

The cells were isolated from mouse lungs using collagenase/dispase digestion followed by filtration with 70 μm Nylon cell strainer and red blood cell lysis; then the cells were incubated with the antibodies against cell surface antigens after blocked with αCD16/CD32. The cells were then fixed with paraformaldehyde (2%), permeabilized and incubated with antibodies against intracellular antigens if needed. For interferon-γ (IFNγ) staining, cells were treated with phorbol 12-myristate 13-acetate (PMA, 50Lng/ml), ionomycin (1LμM), brefeldin A (BFA, 3Lμg/ml) and monensin (2LμM) for 4Lh before they were stained for FACS analysis. Data were acquired and analyzed by Accuri C6 or LSRFortessa I (BD Biosciences) and FlowJo software [29].

Quantitative polymerase chain reaction (qPCR) analysis

The indicated tissues or cells were subjected to DNA or RNA extraction, RNA reverse transcription and real-time PCR using trizol, reverse transcriptase, and Power SYBR Green PCR Master Mix (Thermo Fisher Scientific, Waltham, MA USA) according to manufacturer’s protocol. Primer pairs used for qPCR were listed in Supplementary Table S2.

Microsatellite and gene-specific PCR-based LOH analysis of PDLIM2

Genomic DNAs were isolated from human lung tumors and their matched normal tissues using the PureLink Genomic DNA Purification Kit (Invitrogen, Carlsbad, CA, USA), and subjected to semi-quantitative PCR using the primers specific for the microsatellite markers D8S1786 and D8S1752 that straddle the PDLIM2 genetic locus or the PDLIM2 genetic locus itself. Primer pairs used for the assays were listed in Supplementary Table S2.

Statistical analysis

One-way ANOVA power analysis was used to determine the minimum sample size. Animals were randomly assigned to different treatment groups. Measurements were taken from distinct samples. Student’s t test (two tailed) and one-way ANOVA/Tukey’s or two-way ANOVA/Sidak’s test were used to assess significance of differences between two groups and multiple comparisons, respectively. Pearson’s correlation test was used to assess association between PDLIM2 expression with its promoter methylation or genetic deletion and the overlap between PDLIM2 promoter methylation and genetic deletion. All bars in figures represent means ± standard error of the mean (SEM). The p values are indicated as *p < 0.05, **p < 0.01, ns, not statistically significant, except for those shown in figures. The p valuesL<L0.05 and 0.01 are considered statistically significant and highly statistically significant, respectively. Grubbs’ and ROUT outlier tests were established a priori. There were no exclusions in the analyses.

Results

Both epigenetic repression and genetic deletion of PDLIM2 in lung cancer

Using 40% of the expression level in matched normal lung tissues as the cut-off in the analysis of the Cancer Genome Atlas (TCGA) data, we found that PDLIM2 was repressed in over 75% of human lung tumors in a prior study [6]. If using 50% as the cut-off, PDLIM2 was repressed in about 93% of human lung tumors (Figure 1A). Using 125% of the methylation level of the PDLIM2 promoter in normal lung tissues as the cut-off to analyze TCGA database, over 70% of human lung tumors were found to have hypermethylation of the PDLIM2 promoter (Figure 1A, 1B), further validating our previous finding of epigenetic silencing as the main mechanism underlying PDLIM2 repression [6].

PDLIM2 repression in human lung cancer involves both epigenetic alteration and genetic deletion, and Pdlim2 genetic deletion in mice leads to development of spontaneous tumors, majorly lung tumors. (A) TCGA data showing PDLIM2 repression in over 90% of lung tumors if using 50% of the expression level in normal lung tissues as the cut-off (NL, n = 110; Tumor, n = 1019). (B) TCGA data showing PDLIM2 promoter hypermethylation and expression repression (dashed box) in over 70% of lung tumors when using 125% of the methylation level in normal lung tissues as the cut-off (n = 827). (C) TCGA data showing positive associations between PDLIM2 expression and its gene copy numbers as well as PDLIM2 genetic deletion and expression repression (dashed box) in about 58% of lung tumors using the copy number variation of -0.1 as the cut-off (n = 1010). (D) TCGA data showing simultaneous promoter hypermethylation and genomic deletion of PDLIM2 (dashed box) in about 44% of lung tumors (n = 816). (E) Microsatellite-PCR showing PDLIM2 loss in human lung tumors (n = 21). (F) qPCR showing PDLIM2 loss in human lung cancer cell lines with known copy number of the PDLIM2 gene (n = 25). (G) Kaplan-Meier tumor-free survival curve showing increased spontaneous tumors in Pdlim2-/- and Pdlim2+/- mice compared to WT mice. Gehan-Breslow-Wilcoxon test was performed. *p < 0.05. (H) Percentage of tumor types spontaneously developed in Pdlim2-/- and Pdlim2+/- mice showing a majority of lung tumors.

In line with PDLIM2’s location on chromosome 8p21.3, a frequent LOH region in lung and other tumors [30–34], analysis of TCGA database revealed that PDLIM2 expression was positively associated with its gene copy numbers, and that over 58% of human lung tumors had genetic deletion of the PDLIM2 gene if copy number variation (CNV) of -0.1 was used as the cut-off for the gene deletion (Figure 1C). Further analysis indicated that about 44% of human lung tumors simultaneously harbored the promoter hypermethylation and LOH of the PDLIM2 gene, and approximately 26% and 14% of them only having the promoter hypermethylation or LOH of the PDLIM2 gene, respectively (Figure 1D). Around 16% of lung tumors possessed no such epigenetic or genetic alterations in PDLIM2. These findings were confirmed by microsatellite and gene-specific PCR-based LOH analysis of human primary lung tumor tissues and cell lines (Figure 1E, 1F; Supplementary Table S3). These data suggested that PDLIM2 repression in lung cancer involves both epigenetic silencing and genetic deletion.

To determine the significance of PDLIM2 LOH in lung tumors, we examined whether Pdlim2 heterozygous deletion leads to decreased PDLIM2 expression and spontaneous tumors in mice. Unlike its absolute absence in Pdlim2 homozygous deletion (Pdlim2-/-) mice, PDLIM2 was detected in the lung of Pdlim2+/- mice, but at a much lower level compared to wild type (WT) mice (Supplementary Figure S1A). Importantly, Pdlim2+/- mice, like Pdlim2-/- mice, also developed spontaneous tumors (Figure 1G). Of note, over 50% of tumors developed in Pdlim2-/- and Pdlim2+/- mice were lung tumors (Figure 1H; Supplementary Figure S1B). These data are also highly consistent with the fact that PDLIM2 is ubiquitously expressed under physiological conditions, with the highest level in the lung and lung epithelial cells in particular [6, 18–20]. Thus, PDLIM2 is a haploinsufficient tumor suppressor that is particularly important for lung tumor suppression.

Efficacy of systemic administration of nanoPDLIM2 in refractory lung cancer

Although reversal of PDLIM2 epigenetic repression by epigenetic drugs to restore PDLIM2 expression in cancer cells may be used to treat lung cancer [6], it is logical that this approach cannot be applied to lung tumors involving PDLIM2 LOH, which accounts for about 58% of all lung cancer cases. To overcome this limitation and expand PDLIM2-targeted therapy to all lung tumors with PDLIM2 repression regardless of the involved mechanisms, we tested the therapeutic efficacy of systemic administration of PDLIM2-expression plasmids encapsulated by the clinically feasible in vivo-jetPEI [35–39]. To this end, we employed mouse lung tumors induced by urethane, a faithful model of human lung cancer and adenocarcinoma (AC) in particular [6, 23, 24, 40–43]. Urethane is a chemical carcinogen present in fermented food, alcoholic beverage and also cigarette smoke, the predominant risk factor accounting for about 90% of human lung cancer cases [44]. Like its human counterpart, the murine lung cancer induced by urethane also shares PDLIM2 repression, in addition to their similarities in histology, genetics, molecular biology and immunology [6, 9, 23, 24, 40–43]. WT mice with lung tumors induced by urethane were i.v. injected with nanoparticle-encapsulated Pdlim2 plasmids or empty vector plasmids. Six weeks post the initial treatment of nanotherapy, mice were euthanized and lung tissues were collected (Figure 2A). Compared to the vector plasmid mock-treated group (Vec), nanoPDLIM2-treated mice had a significantly reduced tumor burden in their lungs, although the decrease in tumor numbers was not statistically significant (Figure 2B).

Systemic administration of PDLIM2 plasmid nanoparticles shows efficacy in mouse model of refractory lung cancer. (A) Schedule of lung cancer induction and treatment. (B) Urethane model showing efficacy of intravenous administration of PDLIM2-expression plasmid nanoparticles for refractory lung cancer (n ≥ 6). Nanoparticles with an empty vector plasmid (Vec) that was employed to express PDLIM2 were used as a control. (C) IHC staining showing decreased nuclear expression of STAT3 and RelA in lung tumors by PDLIM2 nanotherapy (n = 6). (D) IHC staining showing decreased Bcl-xL and increased apoptosis marker cleaved caspase -3 in lung tumors by PDLIM2 nanotherapy (n = 6). (E) IHC staining showing decreased Cyclin D1 and proliferation (BrdU incorporation) in lung tumors by PDLIM2 nanotherapy (n = 6). Scale bar in (C-E), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ±LSEM in (B-E). **p < 0.01; ns, not statistically significant.

In line with the tumor reduction, nanoPDLIM2 administration decreased nuclear RelA and STAT3, a hallmark of NF-κB and STAT3 activation, accordingly reduced the expression of their downstream cell survival gene Bcl-xL and cell proliferation gene Cyclin D1, increased apoptosis and decreased proliferation of lung cancer cells (Figure 2C-2E). These data indicated that intravenous administration of nanoPDLIM2 shows efficacy in suppressing oncogenic RelA and STAT3 activation and in treating lung cancer.

High tumor specificity and low toxicity of systemic administration of nanoPDLIM2

To characterize the potential new lung cancer therapy, we examined the expression levels of PDLIM2 in lung tumors and several organs, including the liver, kidney, and spleen. Consistent with the treatment efficacy, a high level of PDLIM2 was detected in the lung tumors from mice treated with the PDLIM2 plasmid nanoparticles one week post nanoPDLIM2 treatment, whereas no obvious PDLIM2 was found in the lung tumors from mice treated with the control plasmid nanoparticles (Figure 3A, 3B). It should be pointed out that the tumor delivery efficiency of PDLIM2-expression or control plasmid nanoparticles was similarly high (Figure 3B). However, either PDLIM2-expression or control plasmids were hardly detected in other organs/tissues of the same mice, including liver, kidney and spleen. Of note, in the first two days after injection, both empty vector and PDLIM2-expression plasmids were also detected at high levels in those tissues but quickly cleared afterward (data not shown), consistent with the well-documented tumor-specific enrichment of nanoparticles [37, 45–47]. Consistently, ectopic Pdlim2 was not detected, and the levels of PDLIM2 proteins were comparable in these tissues (Figure 3B, 3C).

PDLIM2 nanotherapy shows high tumor specificity and low toxicity. (A) IHC staining showing high PDLIM2 re-expression in lung tumors after PDLIM2 nanotherapy. (B) PCR and IB assays showing lung tumor-specific plasmids delivery and PDLIM2 expression by PDLIM2 nanotherapy (n = 3). (C) IHC staining showing comparable expression of PDLIM2 in the indicated tissues of mice treated with PDLIM2 expression plasmid or empty vector plasmid nanoparticles. (D) No significant changes in animal body weight by nanoPDLIM2 (n = 5). (E) H&E staining showing no noticeable changes in major organs by nanoPDLIM2. Scale bar: (A and C) 20 μm; (E) 50 μm.

In further support of the lung tumor-specific delivery, the i.v. injection of either PDLIM2-expression or control plasmid nanoparticles showed no obvious toxicity to animals, as evidenced by no significant changes in the animal body weight and histology of all organs/tissues we examined, including spleen, kidney and liver (Figure 3D, E). Mouse appearance and behaviors, such as eating, drinking, defecating, urinating, sniffing, grooming, and digging, were not different between PDLIM2-expression and control plasmid nanoparticle groups (data not shown). Taken together, these data suggested the therapeutic efficacy and low toxicity of PDLIM2-based nanotherapy in the mouse model of refractory lung cancer.

Synergy of nanoPDLIM2 with chemotherapy in lung cancer treatment

Given the role of PDLIM2 in inhibiting the expression of cell survival and proliferation genes in tumor cells, which contribute to chemoresistance, we tested whether nanoPDLIM2 increases the efficacy of chemotherapy in the mouse model of lung cancer. Treatment with carboplatin and paclitaxel, two chemotherapeutic drugs that are often used together as the first-line treatment for lung and many other cancers, led to significant decrease in tumor number and tumor burden (Figure 4A). Importantly, combination with nanoPDLIM2 further significantly decreased both tumor number and tumor burden, suggesting a promising synergy between PDLIM2 nanotherapy and chemotherapy in lung cancer treatment. Consistently, significantly higher tumor cell apoptosis was detected in mice treated with the combination therapy, in comparison to those with nanoPDLIM2 or chemotherapy alone (Figure 4B).

PDLIM2 nanotherapy renders lung cancers more vulnerable to chemotherapy. (A) Urethane model showing synergy of PDLIM2 nanotherapy and chemotherapy in lung cancer treatment (n ≥ 5). (B) IHC staining showing increased lung tumor cell apoptosis by PDLIM2 nanotherapy, chemotherapy, and further increase by their combination (n = 6). (C) IHC staining showing RelA activation by chemotherapy and blockage of chemo activation of RelA by PDLIM2 nanotherapy (n = 6). (D) IHC staining showing strong MDR1 induction by chemotherapy and blockage of MDR1 induction by PDLIM2 nanotherapy (n = 6). Scale bar in (B-D), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ± SEM. *p < 0.05; **p < 0.01; ns, not statistically significant.

Another important mechanism contributing to the synergy between these two therapies involves nanoPDLIM2 blockade of the acquired chemoresistance of lung cancer cells. Chemotherapy induced strong RelA activation/nuclear expression and MDR1 expression (Figure 4C, 4D). NanoPDLIM2 not only repressed the constitutive activation of RelA in cancer cells but also prevented the strong induction of RelA activation and MDR1 expression by the chemotherapy. Thus, PDLIM2 nanotherapy improves the therapeutic efficacy of chemotherapy through blocking both intrinsic and acquired chemoresistance of lung cancer cells.

NanoPDLIM2 enhancement of ICI’s efficacy in lung cancer treatment

High expression of cell survival genes also renders tumor cells resistant to the tumoricidal activity of cytotoxic T lymphocytes (CTLs), including those unleashed by ICIs. PDLIM2 nanotherapy should also enhance the efficacy of immunotherapy, given its ability in suppressing the expression of cell survival genes. Furthermore, NanoPDLIM2 increased the number of TILs and the expression of MHC-I, the most important and core components of immunotherapies including PD-1 immune checkpoint blockade therapy (Figure 5A, 5B). In line with our previous studies [6], PD-1 blocking antibody showed some efficacies in the mouse model of lung cancer, as evidenced by the significant decrease in tumor burden and tumor number (Figure 5C). Combining with PDLIM2 nanotherapy further significantly decreased tumor burden, although the decrease in tumor number failed to reach statistical significance. Consistently, significant increases in both CD4+ and CD8+ TILs, CD8+ CTL activation and lung tumor cell death were detected (Figure 5D-5F).

PDLIM2 nanotherapy increases the efficacy of PD-1 blockade immunotherapy for refractory lung cancer. (A) IHC staining showing increased TILs in lung tumors by PDLIM2 nanotherapy (n = 6). FACS showing increased MHC-I expression in lung tumor cells by PDLIM2 nanotherapy (n = 4). Urethane model showing PDLIM2 nanotherapy enhancing PD-1 immunotherapy efficacy in lung cancer treatment (n ≥ 5). (D) IHC staining showing increased TILs by PDLIM2 nanotherapy in the context of immunotherapy (n = 6). (E) FACS showing increased activation of CD8+ T cells by PDLIM2 nanotherapy in the context of immunotherapy (n = 3). (F) IHC staining showing increased lung tumor cell apoptosis by PDLIM2 nanotherapy, immunotherapy, and further increase by their combination (n = 6). Scale bar in (A, D, F), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ±LSEM. **p < 0.01; ns, not statistically significant.

Complete remission of all lung tumors in most mice by the combination treatment of nanoPDLIM2 and anti-PD-1 and chemotherapeutic drugs

Like most human lung tumors, lung tumors in our animal model exhibit decreased expression of PD-L1 on cell surface, in additional to the low expression of MHC-I and low number of TILs [6, 9]. In line with our previous finding that PD-L1 expression is largely independent of PDLIM2 [6], nanoPDLIM2 failed to induce PD-L1 expression in lung tumors, which was in contrast to chemotherapy (Figure 6A). This may explain why the enhancing effect of nanoPDLIM2 on PD-1 blockade therapy is only moderate.

Combination of PDLIM2 nanotherapy, chemotherapy and immunotherapy shows great efficacy in lung cancer treatment. (A) IHC staining showing PD-L1 induction by chemotherapy but not PDLIM2 nanotherapy. (B) Schedule of lung cancer induction and treatment. (C) Urethane model showing high resistance of lung tumors to the chemo and αPD-1 combination therapy in lung epithelial specific PDLIM2 deletion mice (ΔSPC) (n ≥ 4). (D) IHC staining showing no MHC-I induction by chemotherapy, PD-1 immunotherapy or their combination (n = 6). (E) Tumor examination showing complete remission of all lung tumors in 60% of mice by combination of the three therapies (n ≥ 5). (F) IHC staining showing increased TILs by PDLIM2 nanotherapy in mice treated with anti-PD-1 and chemotherapeutic drugs (n = 6). (G) FACS analysis showing increased lung CD8+ T-cell activation by PDLIM2 nanotherapy in mice treated with anti-PD-1 and chemotherapeutic drugs (n = 5). (H) No significant effect of PDLIM2 nanotherapy on the body weight of mice treated with anti-PD-1 and chemotherapeutic drugs (n = 5). Scale bar in (A and F), 20 μm. Student’s t test was performed (two tailed, unpaired) and data represent means ±LSEM in (c-g). *p < 0.05; **p < 0.01; ns, not statistically significant.

Although it dramatically increases TILs as well [6], chemotherapy also induced the expression of PD-L1 on cancer cells (Figure 6A), which presumably protects cancer cells from immune attack and thereby restricts further efficacy improvement of its combination with PDLIM2 nanotherapy. But it may increase the sensitivity of cancer cells to PD-1 blockade therapy, particularly in combination with nanoPDLIM2. Indeed, chemotherapy and PD-1 blockade therapy showed a promising synergy in reducing both tumor number and tumor burden in comparison to the individual treatment of chemotherapy or PD-1 blockade therapy [6]. The synergy was largely blocked when Pdlim2 was genetically deleted from lung cancer cells (ΔSPC) (Figure 6B, 6C), suggesting an important role of PDLIM2 in the combination therapy of chemotherapeutic drugs and anti-PD-1. Whereas PDLIM2 nanotherapy increased MHC-I expression on tumor cells (Figure 5B), which is critical for better recognition and killing of tumor cells by CD8+ CTLs, chemotherapy alone or its combination with anti-PD-1 failed to do so (Figure 6D). Accordingly, combination of anti-PD-1 and chemotherapeutic drugs, like the combination therapies of nanoPDLIM2 and anti-PD-1 or chemotherapeutic drugs, also failed to induce a complete remission of lung tumors in any mice.

Given their overlapping roles in increasing TILs and in particular their complement roles in inducing MHC-I and PD-L1 expression on tumor cells, which turn tumors hot and more sensitive to PD-1 blockade, combination of nanoPDLIM2 and chemotherapeutic drugs with anti-PD-1 is expected to show better efficacy compared to combining only two of them. Indeed, combination of all three showed the strongest effect on reducing tumor burden and numbers, and caused complete remission of all lung tumors in 60% of mice (Figure 6E) and much more shrinkage of lung tumors in remaining mice in comparison to those treated with two combined therapies. In line with the high therapeutic efficacy, the triple combination therapy significantly increased the numbers and/or activation of CD4+ and CD8+ T cells in the lung, compared to the combination of chemotherapeutic drugs and anti-PD-1 (Figure 6F, 6G).

Of note, consistent with the undetectable toxicity of PDLIM2 nanotherapy, its co-treatment did not further increase the toxicity of anti-PD-1 and chemotherapeutic drugs. There was no obvious histological difference of major organs, including the liver, lung, kidney, and spleen (Supplementary Figure S2). Moreover, no significant differences in animal body weights were observed by additional nanoPDLIM2, in comparison to the mice received the combinational treatment of anti-PD-1 and chemotherapeutic drugs in the presence or absence of empty vector plasmid nanoparticles (Figure 6H). These data suggested a novel combination therapy with very high therapeutic efficacy and no increased toxicity for lung cancer, particularly refractory lung cancer.

Discussion

PD-1/PD-L1 blockade immunotherapy has recently joined chemotherapy as a standard treatment for lung and several other cancers [2–4]. While some patients have shown dramatic responses, most patients do not benefit from this novel treatment. Currently, various combinations of ICIs with other therapies, in particular its combination with chemotherapeutic drugs are being extensively tested in both preclinical and clinical trial studies to expand the benefit of this innovative therapy [6–8]. Although a promising synergy and better efficacy has been observed in both preclinical animal models and human clinical trial studies [6–8], significant further improvement is direly needed. Using an authentic mouse model of lung cancer, we show, for the first time, that PDLIM2 nanotherapy shows efficacy and high safety, and more importantly, induces complete remission of all lung tumors in most animals when it is combined with anti-PD-1 and chemotherapeutic drugs.

Most human lung tumors as well as lung tumors in our animal model have low numbers of TILs and decreased expression of PD-L1 and MHC-I on the cell surface [6, 9], all of which are important mechanisms contributing to the resistance to PD-1 blockade therapy. Through inducing immunogenic cell death (ICD) of cancer cells [8], chemotherapy can increase TILs and PD-L1 expression on tumor cells, and thereby synergize with anti-PD-1. However, chemotherapy cannot induce MHC-I expression, which limits further improvement of its synergy with immune therapy for complete cancer remission. Also, tumor cells usually express high levels of survival genes against the tumoricidal effects of chemotherapeutic drugs and of CTLs, including those activated by chemotherapy and unleashed by ICIs.

On the other hand, PDLIM2 nanotherapy induces MHC-I expression and lymphocyte tumor infiltration but does not up-regulate PD-L1. Moreover, PDLIM2 nanotherapy prevents the induction of MDR1 and the expression of tumor-related genes and in particular cell survival genes, further sensitizing tumor cells to the cytotoxicity of chemotherapeutic drugs and immune cells including those recruited by chemotherapy and unleashed by PD-1 blockade. Because of these important functions of PDLIM2 nanotherapy, PDLIM2 nanotherapy improved the efficacy of chemotherapy and PD-1 blockade therapy, and in combination with chemotherapy and PD-1 blockade therapy, resulted in complete cancer remission in most of the animals and dramatic tumor reduction in the remaining mice.

Another important clinical characteristic of PDLIM2 nanotherapy is its tumor-specificity and high safety profile. It delivered PDLIM2-expression plasmids to lung tumor tissues and showed undetectable toxicity in the animal model. Its combination does not further increase the toxicity of anti-PD-1 and chemotherapeutic drugs. This is in sharp contrast to the Food and Drug Administration (FDA)-approved epigenetic drugs, which can restore PDLIM2 expression in cancer cells with PDLIM2 epigenetic repression [6, 10–15]. Epigenetic drug treatment leads to body weight loss of animals (Supplementary Figure S3A). The toxicity is much worse when epigenetic drugs are combined with chemotherapy (Supplementary Figure S3).

Besides its side effects, epigenetic therapy should have much more limited application, even though it shows better efficacy as a monotherapy or combined with chemotherapeutic drugs or anti-PD-1 in comparison to PDLIM2 nanotherapy (Supplementary Figure S4). While nanoPDLIM2-based combination therapies could be applicable to all lung tumors with PDLIM2 repression regardless of the mechanisms involved, epigenetic therapy may be used to treat about 26% of lung tumors with PDLIM2 epigenetic repression only. About 58% of lung tumors harboring PDLIM2 LOH are not suitable to epigenetic therapies, although most of them are also with epigenetic alterations of the pdlim2 gene. Of note, the therapeutic efficacy of epigenetic therapy depends on PDLIM2 expression [6], and PDLIM2 heterozygous loss causes spontaneous lung and other cancers. Further, the efficacy and toxicity of the combination of epigenetic agents with both anti-PD-1 and chemotherapeutic drugs has not been examined yet.

Conclusions

In summary, the presented data identify genetic deletions as a major mechanism other than epigenetic alterations for PDLIM2 repression in human lung cancer, and PDLIM2 as a haploinsufficient tumor suppressor particularly important for suppressing lung cancer and therapy resistance. More importantly, these preclinical data establish a novel combination treatment of nanoPDLIM2, anti-PD-1 and chemotherapeutic drugs that induces complete remission of all lung tumors in most animals and is also with high safety profile. We believe that these knowledges are applicable to other cancers, because PDLIM2 repression has also been linked to numerous human cancers other than lung cancer.